The imprint of global magma oceans on exoplanet demographics

### 2023-04-14 src/tex/ms.tex
reframe "assumptions" sections from caveats to future refinement

\subsection{Impact of assumptions on our findings}
\todo[inline]{Caveats\\
- not yet explored: how sensitive are we on the assumed planet population? $\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert? Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}. We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
The prospects for detecting the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.

\subsubsection{Structures in the planet occurrence rate density}
In principle, the baseline occurrence rate density in radius-period space that governs the generation of our synthetic planet sample might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
We thus do not expect the model underlying our planet sample to affect the investigation of the habitable zone inner edge discontinuity.
If any currently unknown sharp features in the distribution of terrestrial planets emerge, they should be considered in future studies.
%However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may change the magnitude of radius inflation and could introduce an offset to its demographic imprint, we do not expect a change of its structure.
The magnitude of the radius inflation effect is in fact expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep[][, add Noack+ paper]{Zeng2016,2021JGRE..12606724B}. CITE!
We thus consider our mass-radius relationship a conservative case and do not expect a negative impact on the detectability of the runaway greenhouse transition due to deviations from it.
Nevertheless, we performed a sanity check to assess the influence varying our baseline model on our results (see Appendix~\ref{app:MR_relation}), and found general agreement between different interior composition models.

The models in this study do not take into account a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019}, and we neglect any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
Future theoretical work that self-consistently models interior-atmosphere interactions is needed to take these effects into account~\citep{2022arXiv220310023L,Wordsworth2022}.

\subsubsection{Simplifications in the atmosphere model}
%Mention we only consider instellation-induced rgh.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
%}
% >>>>>>>>>>
% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv. Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted. I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
% >>>>>>>>>>
As illustrated in Figure~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community~\citep{2021SSRv..217...22G,2022arXiv220310023L,Wordsworth2022}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
The real counterpart to our idealized step-function central to our hypothesis tests could be blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the habitable zone inner edge discontinuity, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.

Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
The choice of statistical estimator for the hypothesis tests may further blur the recovered discontinuity.
We compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor'' we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.

>>>>

\subsection{Impact of assumptions on our findings}
%\todo[inline]{Caveats\\
%- not yet explored: how sensitive are we on the assumed planet population? $\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert? Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}. We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
The prospects for probing the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
%We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
Our state-of-the-art models approximate the situation and offer testable predictions.
In the following, we review a few considerations that future models may include to refine these predictions.

\subsubsection{Structures in the planet occurrence rate density}
The baseline occurrence rate density in radius-period space that governs the generation of synthetic planets might influence our findings, especially if it contains any features that coincide with the injected demographic feature.
This is not the case in the model from \citet{Bergsten2022} that we adopted: Its occurrence rate density varies smoothly in the domain relevant for the runaway greenhouse hypothesis; transitions only occur at smaller instellations ($< \SI{50}{\watt\per\meter\squared}$) and larger radii ($> 1.6\,R_\oplus$).
We thus do not expect the model underlying our planet sample to affect the investigation of the habitable zone inner edge discontinuity.
If any currently unknown sharp features in the distribution of terrestrial planets emerge, they should be considered in future studies.
%However, it should be kept in mind that any unknown sharp features in the true distribution of terrestrial planets could complicate a detection or lead to a false positive result (see Sect.~\ref{sec:dis_falsepositive}).

\subsubsection{Baseline mass-radius relationship}
Our baseline mass-radius relationship assuming pure $\mathrm{MgSiO_3}$ interiors~\citep{Zeng2016} might not be representative of the rocky planet population.
However, while interior composition may introduce an offset to the radius habitable zone inner edge discontinuity, we do not expect a change of its structure.
Since the magnitude of the radius inflation effect is expected to be larger for an Earth-like interior composition with an iron core-silicate mantle structure~\citep{Zeng2016,2021JGRE..12606724B},
\todo[inline]{Tim: add Noack+ paper}
we consider our mass-radius relationship a conservative case.
We performed a sanity check to assess the impact of varying our baseline model (see Appendix~\ref{app:MR_relation}) and found general agreement between different interior compositions.

The models in this study neglect a predicted additional radius increase on the percent level due to a molten interior~\citep{Bower2019} and any potential effects stemming from a deviating gas exchange between atmosphere and interior in super-runaway greenhouse planets due to different redox conditions~\citep{Ikoma2018,2021JGRE..12606711L,2022PSJ.....3...93B,2021SSRv..217...22G}.
Future self-consistent modeling of interior-atmosphere interactions may fill these gaps~\citep{2022arXiv220310023L,Wordsworth2022}.

\subsubsection{Atmosphere model}
%Mention we only consider instellation-induced rgh.\\
%- we ignore water loss by H$_2$O photolysis in the upper atmosphere and subsequent H escape, which would eventually cool even planets %within the runaway greenhouse regime~\citep{2022arXiv220310023L}.\\
%- we ignore tidal heating, which could extend the magma ocean phase of close-in planets (and change their orbits via tidal orbital evolution)( Jackson et al., 2008; Barnes et al., 2013 (?)).\\
%- assumption of a single albedo. ``if there is a range of albedos, will that wash away the signal?''
%}
% >>>>>>>>>>
% TL: I just wrote a review article about this exact problem (magma ocean atmosphere exchange) with Yamila Miguel, hopefully soon on arXiv. Honestly, I think we keep it short here, otherwise the paper will get way too long and diluted. I referenced the latest reviews on this now, and we can cite Lichtenberg & Miguel 23 as soon as it's on arXiv for a much much longer discussion.
% >>>>>>>>>>
As illustrated in Figure~\ref{fig:radiusevolution}, evolutionary processes and a background population of planets unable to sustain a steam atmosphere complicate the classification of a planet into the super- or sub-runaway category.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a planet in our runaway greenhouse category, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to the sub-runaway greenhouse category on a timescale of $\SI{e8}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
We marginalized over these degeneracies by introducing the dilution factor $f_\mathrm{rgh}$, but in a real survey contextual information about a planet's age or likely formation history could be taken into account.
Developing more sophisticated models of the coupled interaction between atmosphere and interior for an arbitrary planetary composition remains a long-term goal of the community~\citep{2021SSRv..217...22G,2022arXiv220310023L,Wordsworth2022}.

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity. %, and the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
The real counterpart to the idealized step-function in our hypothesis tests could be blurred by several factors.
Planets that lack an atmosphere or sufficient volatiles cannot bear a steam atmosphere.
Those with significant Hydrogen/Helium envelopes can disguise as inflated rocky planets and might not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was recently suggested for water-dominated ``sub-Neptunes''~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
A time dependence is introduced through stellar luminosity evolution and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.

Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Turbet2021,Pierrehumbert2022}.
The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

All of the above mechanisms may affect the habitable zone inner edge discontinuity, and their contributions are challenging to quantify.
However, while they may offset the signal's amplitude, they preserve its general shape.
The ``dilution factor'' we introduced embodies our ignorance of the magnitude of this offset.
